MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS

### 2024-12-30 manuscript/main.tex
:construction: Add interference channel results

	% in two common scenarios: 1) forward/backward channel is rank-deficient and 2) direct channel is blocked.

@@

	\begin{subsection}{Related Works}
		At a specific resource block, channel shaping metrics can be classified into two categories:
		% shaping metrics can be classified into two categories:
		\begin{itemize}
			\item \emph{Singular value:} The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}. Those are closely related to performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to minor perturbations of the channel matrix.
			\item \emph{Power:} The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} in point-to-point channels and leakage interference \cite{Santamaria2023a} in interference channels. Those second-order metrics are less informative in \gls{mimo} but easier to analyze and optimize.
		\end{itemize}

@@

		Fourth, we provide an analytical answer to the shaping question in typical deployment scenarios.
		% For rank-deficient forward/backward channel,
		When the forward/backward channel is rank-deficient, we derive singular value bounds applying to \gls{d}- and \gls{bd}-\gls{ris} with asymptotically large number of elements.
		% For negligible direct channel
		When the direct channel is negligible, we derive singular value bounds applying to fully-connected \gls{bd}-\gls{ris} with arbitrary number of elements.
		Those bounds are validated by comparing with the numerical results above.
		Results show that for a fixed number of elements, \gls{bd}-\gls{ris} can approach the asymptotic bounds better than \gls{d}-\gls{ris}.

@@

		% 1) rank-deficient forward/backward channel and 2) direct channel being blocked.

		% % leverage matrix analysis tools and provide an analytical answer to the singular value shaping question.
		% % revisit the channel singular value shaping problem from an analytical perspective via matrix analysis.
		% % To explore the ultimate shaping limits, we derive individual and collective singular value bounds for rank-deficient forward/backward channel and blocked direct channel.
		% % Asymptotic bounds shared by diagonal and \gls{bd}-\gls{ris} are derived in the former case, while exact bounds for unitary \gls{bd}-\gls{ris} are derived in the latter case.
		% Asymptotic channel singular value bounds applying to diagonal and \gls{bd}-\gls{ris} are derived for rank-deficient forward/backward channel.
		% On the other hand, tight bounds for fully-connected \gls{bd}-\gls{ris} with finite number of elements are derived for blocked direct channel.
		% Results validate those bounds and emphasizes the shaping advantage of \gls{bd}-\gls{ris}.
		% This is the first work to quantify the singular value redistribution capability of \gls{bd}-\gls{ris} from either numerical or analytical perspective.
		% Results validate those bounds and show that
		% fully-connected \gls{bd}-\gls{ris} without direct link.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	% $\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$, $\mathbb{H}_+^{n \times n}$, and $\mathbb{U}^{n \times n}$ denote the set of $n \times n$ Hermitian, positive semi-definite, and unitary matrices, respectively.
	% $\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	% $\tr(\cdot)$ and $\det(\cdot)$ evaluate the trace and determinant of a square matrix, respectively.
	$\diag(\cdot)$ constructs a square matrix with arguments on the main (block) diagonal and zeros elsewhere.
	$\sv(\cdot)$, $\ran(\cdot)$, and $\ker(\cdot)$ evaluate the singular values (in descending order), column space (range), and kernel of a matrix, respectively.
	% $\sv(\cdot)$ extracts the singular values of a matrix as a non-ascending vector.
	% $\card(\cdot)$ returns the cardinality of a set.
	$\conv(\cdot)$ returns the convex hull of arguments.
	$\lvert \cdot \rvert$, $\lVert \cdot \rVert$, and $\lVert \cdot \rVert _\mathrm{F}$ denote the absolute value, Euclidean norm, and Frobenius norm, respectively.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ are the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, and stationary point, respectively.
	$[N]$ is a shortcut for $\{1,2,\ldots,N\}$.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\odot$ denotes the element-wise (Hadamard) product.
	$\mathcal{N}_{\mathbb{C}}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

		\begin{corollary}[Channel power gain]
			\label{co:nd_power}
			If the direct channel is negligible, then the channel power gain is bounded from above (resp. below) by the inner product of squared singular values of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ when they are sorted similarly (resp. oppositely), that is,
			\begin{equation}
				\label{iq:power_nd}
				\sum_{n=1}^N \sigma_n^2(\mathbf{H}_\mathrm{B}) \sigma_{N_\mathrm{S}-n+1}^2(\mathbf{H}_\mathrm{F}) \le \lVert \mathbf{H} \rVert _\mathrm{F}^2 \le \sum_{n=1}^N \sigma_n^2(\mathbf{H}_\mathrm{B}) \sigma_n^2(\mathbf{H}_\mathrm{F}),
			\end{equation}
			whose\footnote{As a side note, we notice \cite{Fang1994} discussed a similar bound using extreme singular values $\max\bigl(\sigma_{N}(\mathbf{H}_\mathrm{B}) \lVert \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2, \sigma_{N}(\mathbf{H}_\mathrm{F}) \lVert \mathbf{H}_\mathrm{B} \rVert _\mathrm{F}^2\bigr) \le \lVert \mathbf{H} \rVert _\mathrm{F}^2 \le \min\bigl(\sigma_1(\mathbf{H}_\mathrm{B}) \lVert \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2, \sigma_1(\mathbf{H}_\mathrm{F}) \lVert \mathbf{H}_\mathrm{B} \rVert _\mathrm{F}^2\bigr)$. This is a looser version of \eqref{iq:power_nd} and cannot take equalities unless the extreme singular values are of multiplicity $N$.} upper and lower bounds are attained respectively at
			\begin{subequations}
				\begin{align}
					\mathbf{\Theta}_\textnormal{P-max}^\textnormal{MIMO-ND} & = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}, \label{eq:ris_nd_power_max}            \\
					\mathbf{\Theta}_\textnormal{P-min}^\textnormal{MIMO-ND} & = \mathbf{V}_\mathrm{B} \mathbf{J} \mathbf{U}_\mathrm{F}^\mathsf{H} \label{eq:ris_nd_power_min},
				\end{align}
			\end{subequations}
			where $\mathbf{J}$ is the exchange (a.k.a. backward identity) matrix of dimension $N_\mathrm{S}$.
		\end{corollary}
		\begin{proof}
			Please refer to Appendix~\ref{ap:nd_power}.
		\end{proof}

@@

		Corollary \ref{co:nd_capacity_snr_general} also suggests that the power- and rate-optimal scattering matrices \eqref{eq:ris_nd_power_max} and \eqref{eq:ris_nd_rate_max} coincide with each other when the direct channel is negligible and the \gls{bd}-\gls{ris} is fully-connected.
		When either condition is not satisfied, active and passive beamforming are coupled and the rate-optimal solution involves alternating optimization.
		However, the power-optimal \gls{ris} still provides for a low-complexity decoupled solution.
		The details will be discussed in Section~\ref{sc:rate_pc}.

@@

		The ergodic counterparts of \eqref{eq:capacity_nd} and \eqref{iq:capacity_nd_snr_extreme} when both $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ follow Rayleigh fading can be evaluated similarly to \eqref{eq:power_nd_rayleigh}.
		Proposition \ref{pp:dof}--\ref{pp:nd} and the resulting Corollaries provide a partial answer to the channel shaping question in terms of singular values and their functions.
		Extending the analysis to more general cases (e.g., non-negligible direct channel and arbitrary \gls{bd}-\gls{ris} group size) is non-trivial due to limited branch matching and mode alignment capabilities therein.
		A numerical solution will be discussed in Section~\ref{sc:shaping_numerical}.
	\end{subsection}

@@

		With Proposition \ref{pp:shaping}, one can apply the \emph{relax-then-project} method \cite{Shen2020a,Fang2023} or \emph{non-geodesic\footnote{A geodesic is a curve representing the shortest path between two points in a Riemannian manifold, whose tangent vectors remain parallel when transporting along the curve.} \gls{rcg}} \cite{Li2023b,Li2023c,Zhou2023} to solve Problem \eqref{op:shaping}.

@@

		\begin{enumerate}

@@

	\begin{section}{Rate Maximization}
		\label{sc:rate}
		In this section, we first solve rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} point-to-point and interference channels optimally by joint beamforming design, and then exploit channel shaping for a low-complexity two-stage solution.

		\begin{subsection}{MIMO Point-to-Point Channel}
			\label{sc:rate_pc}
			The achievable rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} point-to-point channel is formulated as
			\begin{maxi!}
				{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{R = \log \det \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\eta}\biggr)}{\label{op:rate}}{\label{ob:rate}}
				\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2}{\le P}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g}{=\mathbf{I}, \quad \forall g,\label{cn:rate_unitary}}
			\end{maxi!}
			where $\mathbf{W}$ is the transmit precoder, $R$ is the achievable rate, $\eta$ is the average noise power, and $P$ is maximum average transmit power.
			Problem \eqref{op:rate} is non-convex due to the block-unitary constraint \eqref{cn:rate_unitary} and the coupling between variables.
			Two approaches are proposed to solve it optimally or efficiently.
			% We propose a local-optimal approach based on \gls{ao} and a low-complexity approach based on channel shaping.

			\begin{subsubsection}{Alternating Optimization}
				This approach updates $\mathbf{\Theta}$ and $\mathbf{W}$ iteratively until convergence.
				For a given $\mathbf{W}$, the passive beamforming subproblem is
				\begin{maxi!}
					{\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{H} \mathbf{Q} \mathbf{H}^\mathsf{H}}{\eta}\biggr)}{\label{op:rate_ris}}{\label{ob:rate_ris}}
					\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
				\end{maxi!}
				where $\mathbf{Q} \triangleq \mathbf{W} \mathbf{W}^\mathsf{H}$ is the transmit covariance matrix.
				Problem \eqref{op:rate_ris} can be solved optimally by Algorithm \ref{ag:rcg} with the partial derivative given in Lemma \ref{lm:rate}.
				\begin{lemma}
					\label{lm:rate}
					The partial derivative of \eqref{ob:rate_ris} with respect to \gls{bd}-\gls{ris} block $g$ is
					\begin{equation}
						\label{eq:gradient_eucl_rate}
						\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\eta} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \biggl(\mathbf{I} + \frac{\mathbf{H}\mathbf{Q}\mathbf{H}^\mathsf{H}}{\eta}\biggr)^{-1} \mathbf{H} \mathbf{Q} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
					\end{equation}
				\end{lemma}

@@

			\begin{subsubsection}{Low-Complexity Solution}
				\label{sc:rate_pc_lc}
				To reduce the computational complexity, we suboptimally decouple the beamforming design by first shape the channel by \gls{ris} for maximum power gain and then optimize the active beamforming.
				The channel power gain maximization problem is formulated as\footnote{Problem \eqref{op:power} has been studied in \gls{siso} \cite{Shen2020a} and \gls{miso} equivalents \cite{Santamaria2023,Fang2023,Nerini2023,Nerini2023b} where only one mode is available. Generalizing those to \gls{mimo} is non-trivial due to trade-off between modes.}
				\begin{maxi!}
					{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power}}{\label{ob:power}}
					\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{\label{cn:power_unitary}}{}
				\end{maxi!}
				Inspired by \cite{Nie2017}, we propose a closed-form iterative solution that converges to a stationary point of Problem \eqref{op:power}.
				The idea is to approximate the quadratic objective \eqref{ob:power} successively by Taylor expansion and solve each subproblem by group-wise \gls{svd}.

@@

		\begin{subsection}{MIMO Interference Channel}
			On top of \eqref{eq:channel_interference}, the achievable rate of transmission $k$ is
			\begin{equation}
				R_k = \log \det \biggl(\mathbf{I} + \mathbf{W}_k {\mathbf{H}^{(kj)\mathsf{H}}} \mathbf{Q}_k^{-1} {\mathbf{H}^{(kj)}} \mathbf{W}_k\biggr),
			\end{equation}
			where $\mathbf{W}_k$ is the precoder at transmitter $k$ and $\mathbf{Q}_k = \sum_{j \ne k} {\mathbf{H}^{(kj)}} \mathbf{W}_j \mathbf{W}_j^\mathsf{H} {\mathbf{H}^{(kj)\mathsf{H}}} + \eta \mathbf{I}$ is the interference-plus-noise covariance matrix at receiver $k$.
			The weighted sum-rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} interference channel is formulated as
			\begin{maxi!}
				{\scriptstyle{\mathbf{\Theta}, \{\mathbf{W}_k\}_{k \in [K]}}}{\sum_{k=1}^K \rho_k R_k}{\label{op:wsr}}{\label{ob:wsr}}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g}{}{}
				\addConstraint{\lVert \mathbf{W}_k \rVert _\mathrm{F}^2 \le P_k. \quad \forall k}{}{}
			\end{maxi!}
			where $\rho_k \ge 0$ is the weight associated with transmission $k$.
			This non-convex problem can be solved by extending both solutions covered in Section \ref{sc:rate_pc} as detailed below.

			\begin{subsubsection}{Alternating Optimization}
				This approach updates $\mathbf{\Theta}$ and $\{\mathbf{W}_k\}_{k \in [K]}$ iteratively until convergence.
				For a given precoder set, the passive beamforming subproblem is
				\begin{maxi!}
					{\scriptstyle{\mathbf{\Theta}}}{\sum_{k=1}^K \rho_k R_k}{\label{op:wsr_ris}}{\label{ob:wsr_ris}}
					\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
				\end{maxi!}
				which can be solved optimally by Algorithm \ref{ag:rcg} with the partial derivative given in Lemma \ref{lm:wsr}.

>>>>

	% in two common scenarios: 1) backward/forward channel is rank-deficient and 2) direct channel is blocked.

@@

	\begin{subsection}{Related Works}
		At a specific resource block, channel shaping metrics can be categorized into two types.
		% shaping metrics can be classified into two categories:
		\begin{itemize}
			\item \emph{Singular value:} The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}. Those are closely related to performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to minor perturbations of the channel matrix.
			\item \emph{Power:} The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} in point-to-point channels and leakage interference \cite{Santamaria2023a} in interference channels. Those second-order metrics are less informative in \gls{mimo} but easier to analyze and optimize.
		\end{itemize}

@@

		Fourth, we provide an analytical answer to the shaping question in typical deployment scenarios.
		% For rank-deficient backward/forward channel,
		When the backward/forward channel is rank-deficient, we derive singular value bounds applying to \gls{d}- and \gls{bd}-\gls{ris} with asymptotically large number of elements.
		% For negligible direct channel
		When the direct channel is negligible, we derive singular value bounds applying to fully-connected \gls{bd}-\gls{ris} with arbitrary number of elements.
		Those bounds are validated by comparing with the numerical results above.
		Results show that for a fixed number of elements, \gls{bd}-\gls{ris} can approach the asymptotic bounds better than \gls{d}-\gls{ris}.

@@

		% 1) rank-deficient backward/forward channel and 2) direct channel being blocked.

		% % leverage matrix analysis tools and provide an analytical answer to the singular value shaping question.
		% % revisit the channel singular value shaping problem from an analytical perspective via matrix analysis.
		% % To explore the ultimate shaping limits, we derive individual and collective singular value bounds for rank-deficient backward/forward channel and blocked direct channel.
		% % Asymptotic bounds shared by diagonal and \gls{bd}-\gls{ris} are derived in the former case, while exact bounds for unitary \gls{bd}-\gls{ris} are derived in the latter case.
		% Asymptotic channel singular value bounds applying to diagonal and \gls{bd}-\gls{ris} are derived for rank-deficient backward/forward channel.
		% On the other hand, tight bounds for fully-connected \gls{bd}-\gls{ris} with finite number of elements are derived for blocked direct channel.
		% Results validate those bounds and emphasizes the shaping advantage of \gls{bd}-\gls{ris}.
		% This is the first work to quantify the singular value redistribution capability of \gls{bd}-\gls{ris} from either numerical or analytical perspective.
		% Results validate those bounds and show that
		% fully-connected \gls{bd}-\gls{ris} without direct link.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	% $\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$, $\mathbb{H}_+^{n \times n}$, and $\mathbb{U}^{n \times n}$ denote the set of $n \times n$ Hermitian, positive semi-definite, and unitary matrices, respectively.
	% $\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	% $\tr(\cdot)$ and $\det(\cdot)$ evaluate the trace and determinant of a square matrix, respectively.
	$\diag(\cdot)$ constructs a square matrix with arguments on the main (block) diagonal and zeros elsewhere.
	$\sv(\cdot)$, $\ran(\cdot)$, and $\ker(\cdot)$ evaluate the singular values (in descending order), column space (range), and kernel of a matrix, respectively.
	% $\sv(\cdot)$ extracts the singular values of a matrix as a non-ascending vector.
	% $\card(\cdot)$ returns the cardinality of a set.
	$\conv(\cdot)$ returns the convex hull of arguments.
	$\mathrm{vec}(\cdot)$ stacks the columns of a matrix sequentially into a vector.
	$\lvert \cdot \rvert$, $\lVert \cdot \rVert$, and $\lVert \cdot \rVert _\mathrm{F}$ denote the absolute value, Euclidean norm, and Frobenius norm, respectively.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ are the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, and stationary point, respectively.
	$[N]$ is a shortcut for $\{1,2,\ldots,N\}$.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\odot$ denotes the element-wise (Hadamard) product.
	$\mathcal{N}_{\mathbb{C}}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

		\begin{corollary}[Channel power gain]
			\label{co:nd_power}
			If the direct channel is negligible, then the channel power gain is bounded from above (resp. below) by the inner product of squared singular values of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ when they are sorted similarly (resp. oppositely), that is,
			\begin{equation}
				\label{iq:power_nd}
				\sum_{n=1}^N \sigma_n^2(\mathbf{H}_\mathrm{B}) \sigma_{N_\mathrm{S}-n+1}^2(\mathbf{H}_\mathrm{F}) \le \lVert \mathbf{H} \rVert _\mathrm{F}^2 \le \sum_{n=1}^N \sigma_n^2(\mathbf{H}_\mathrm{B}) \sigma_n^2(\mathbf{H}_\mathrm{F}),
			\end{equation}
			whose\footnote{As a side note, we notice \cite{Fang1994} discussed a similar bound using extreme singular values $\max\bigl(\sigma_{N}(\mathbf{H}_\mathrm{B}) \lVert \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2, \sigma_{N}(\mathbf{H}_\mathrm{F}) \lVert \mathbf{H}_\mathrm{B} \rVert _\mathrm{F}^2\bigr) \le \lVert \mathbf{H} \rVert _\mathrm{F}^2 \le \min\bigl(\sigma_1(\mathbf{H}_\mathrm{B}) \lVert \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2, \sigma_1(\mathbf{H}_\mathrm{F}) \lVert \mathbf{H}_\mathrm{B} \rVert _\mathrm{F}^2\bigr)$. This is a looser version of \eqref{iq:power_nd} and cannot take equalities unless the extreme singular values are of multiplicity $N$.} upper and lower bounds are attained respectively at
			\begin{subequations}
				\label{eq:ris_nd_power}
				\begin{align}
					\mathbf{\Theta}_\textnormal{P-max}^\textnormal{MIMO-ND} & = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}, \label{eq:ris_nd_power_max}            \\
					\mathbf{\Theta}_\textnormal{P-min}^\textnormal{MIMO-ND} & = \mathbf{V}_\mathrm{B} \mathbf{J} \mathbf{U}_\mathrm{F}^\mathsf{H} \label{eq:ris_nd_power_min},
				\end{align}
			\end{subequations}
			where $\mathbf{J}$ is the exchange (a.k.a. backward identity) matrix of dimension $N_\mathrm{S}$.
		\end{corollary}
		\begin{proof}
			Please refer to Appendix~\ref{ap:nd_power}.
		\end{proof}

@@

		Corollary \ref{co:nd_capacity_snr_general} also suggests that the power- and rate-optimal scattering matrices \eqref{eq:ris_nd_power_max} and \eqref{eq:ris_nd_rate_max} coincide with each other when the direct channel is negligible and the \gls{bd}-\gls{ris} is fully-connected.
		When either condition is not satisfied, active and passive beamforming are coupled and the rate-optimal solution involves alternating optimization.
		However, the power-optimal \gls{ris} still provides for a low-complexity decoupled solution.
		The details will be discussed in Section~\ref{sc:rate}.

@@

		The ergodic counterparts of \eqref{eq:capacity_nd} and \eqref{iq:capacity_nd_snr_extreme} when both $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ follow Rayleigh fading can be evaluated similarly to \eqref{eq:power_nd_rayleigh}.
		Proposition \ref{pp:dof}--\ref{pp:nd} and the resulting Corollaries provide a partial answer to the channel shaping question in terms of singular values and their functions.
		Extending the analysis to more general cases (e.g., significant direct channel and arbitrary \gls{bd}-\gls{ris} group size) is non-trivial due to limited branch matching and mode alignment capabilities therein.
		A numerical solution will be discussed in Section~\ref{sc:shaping_numerical}.
	\end{subsection}

@@

		With Proposition \ref{pp:shaping}, one can apply the \emph{relax-then-project} method \cite{Shen2020a,Fang2023} or \emph{non-geodesic\footnote{A geodesic is a curve representing the shortest path between two points in a Riemannian manifold, whose tangent vectors remain parallel when transporting along the curve.} \gls{rcg}} \cite{Li2023b,Li2023c,Zhou2023} to solve problem \eqref{op:shaping}.

@@

		\begin{enumerate}[label=(\roman*)]

@@

\begin{section}{Rate Maximization}
	In this section, we first solve rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} point-to-point and interference channels optimally by joint beamforming design, and then exploit channel shaping for a low-complexity two-stage solution.

	\begin{subsection}{MIMO Point-to-Point Channel}
		\label{sc:rate}
		The achievable rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} point-to-point channel is formulated as
		\begin{maxi!}
			{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{R = \log \det \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\eta}\biggr)}{\label{op:rate}}{\label{ob:rate}}
			\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2}{\le P}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g}{=\mathbf{I}, \quad \forall g,\label{cn:rate_unitary}}
		\end{maxi!}
		where $\mathbf{W}$ is the transmit precoder, $R$ is the achievable rate, $\eta$ is the average noise power, and $P$ is maximum average transmit power.
		Problem \eqref{op:rate} is non-convex due to the block-unitary constraint \eqref{cn:rate_unitary} and the coupling between variables.
		Two approaches are proposed to solve it optimally or efficiently.

		\begin{subsubsection}{Alternating Optimization}
			\label{sc:rate_ao}
			This approach updates $\mathbf{\Theta}$ and $\mathbf{W}$ iteratively until convergence.
			For a given $\mathbf{W}$, the passive beamforming subproblem is
			\begin{maxi!}
				{\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{H} \mathbf{Q} \mathbf{H}^\mathsf{H}}{\eta}\biggr)}{\label{op:rate_ris}}{\label{ob:rate_ris}}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
			\end{maxi!}
			where $\mathbf{Q} \triangleq \mathbf{W} \mathbf{W}^\mathsf{H}$ is the transmit covariance matrix.
			Problem \eqref{op:rate_ris} can be solved optimally by Algorithm \ref{ag:rcg} with the partial derivative given in Lemma \ref{lm:rate}.
			\begin{lemma}
				\label{lm:rate}
				The partial derivative of \eqref{ob:rate_ris} with respect to \gls{bd}-\gls{ris} block $g$ is
				\begin{equation}
					\label{eq:gradient_eucl_rate}
					\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\eta} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \biggl(\mathbf{I} + \frac{\mathbf{H}\mathbf{Q}\mathbf{H}^\mathsf{H}}{\eta}\biggr)^{-1} \mathbf{H} \mathbf{Q} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\end{equation}
			\end{lemma}

@@

		\begin{subsubsection}{Low-Complexity Solution}
			\label{sc:rate_lc}
			To reduce the computational complexity, we suboptimally decouple the beamforming design by first shape the channel by \gls{ris} for maximum power gain and then optimize the active beamforming.
			The channel power gain maximization problem is formulated as\footnote{Problem \eqref{op:power} has been studied in \gls{siso} \cite{Shen2020a} and \gls{miso} equivalents \cite{Santamaria2023,Fang2023,Nerini2023,Nerini2023b} where only one mode is available. Generalizing those to \gls{mimo} is non-trivial due to trade-off between modes.}
			\begin{maxi!}
				{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power}}{\label{ob:power}}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{\label{cn:power_unitary}}{}
			\end{maxi!}
			Inspired by \cite{Nie2017}, we propose a closed-form iterative solution that converges to a stationary point of problem \eqref{op:power}.
			The idea is to approximate the quadratic objective \eqref{ob:power} successively by Taylor expansion and solve each subproblem by group-wise \gls{svd}.

@@

	\begin{subsection}{MIMO Interference Channel}
		On top of \eqref{eq:channel_interference}, the achievable rate of transmission $k$ is
		\begin{equation}
			R_k = \log \det \biggl(\mathbf{I} + \mathbf{W}_k {\mathbf{H}^{(kj)\mathsf{H}}} \mathbf{Q}_k^{-1} {\mathbf{H}^{(kj)}} \mathbf{W}_k\biggr),
		\end{equation}
		where $\mathbf{W}_k$ is the precoder at transmitter $k$ and $\mathbf{Q}_k = \sum_{j \ne k} {\mathbf{H}^{(kj)}} \mathbf{W}_j \mathbf{W}_j^\mathsf{H} {\mathbf{H}^{(kj)\mathsf{H}}} + \eta \mathbf{I}$ is the interference-plus-noise covariance matrix at receiver $k$.
		The \gls{wsr} maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} interference channel is formulated as
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}, \{\mathbf{W}_k\}_{k \in [K]}}}{\sum_{k=1}^K \rho_k R_k}{\label{op:wsr}}{\label{ob:wsr}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g}{}{}
			\addConstraint{\lVert \mathbf{W}_k \rVert _\mathrm{F}^2 \le P_k. \quad \forall k}{}{}
		\end{maxi!}
		where $\rho_k \ge 0$ is the weight associated with transmission $k$.
		This non-convex problem can be solved by extending both solutions covered in Section \ref{sc:rate} as detailed below.

		\begin{subsubsection}{Alternating Optimization}
			\label{sc:wsr_ao}
			This approach updates $\mathbf{\Theta}$ and $\{\mathbf{W}_k\}_{k \in [K]}$ iteratively until convergence.
			For a given precoder set, the passive beamforming subproblem is
			\begin{maxi!}
				{\scriptstyle{\mathbf{\Theta}}}{\sum_{k=1}^K \rho_k R_k}{\label{op:wsr_ris}}{\label{ob:wsr_ris}}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
			\end{maxi!}
			which can be solved optimally by Algorithm \ref{ag:rcg} with the partial derivative given in Lemma \ref{lm:wsr}.
